---
title: "Logistic regression 2"
output: 
  learnr::tutorial:
     allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library('EP03logistic')
data("bmt", package = "EP03logistic")
knitr::opts_chunk$set(echo = FALSE)
```


## Quiz

The following questions test your knowledge about logistic regression, odds and odds ratios.

### Question 1
```{r Q1}
quiz(
  question("True or false (check all that apply)", 
    type='multiple',         
    answer("In regression an outlier is a value with an unsual value of the dependent variable given the independent  variables", correct=TRUE),
    answer("Observations with high leverage have (an) unusual independant variable.", correct=TRUE ),
    answer("An odds of one correspons to a probability of 50%", correct=TRUE),
    answer("The deviance is a measure of leverage"),
    answer("Cook's distance says something about how influencial an observation is", correct=TRUE),
    answer("The dfbetas say something about how influencial an observation is", correct=TRUE),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 2

In Clements et. al. (1999) the effect of a high dose vs a low dose of D-Penicillamine in early diffuse systemic scleloris is investigated. 25 of the 36 patients that received a low dose responded while in the high dose group 17 out of 32 responded.
```{r Q2}
quiz(
  question("What is the probability of responding in the low dose group?",
    answer("Less than 40%"),
    answer("Between 40% and 60%"),
    answer("Between 60% and 80%", correct=TRUE),
    answer("More than 80%"),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 3


```{r Q3}
quiz(
  question("What are the odds of responding in the low dose group?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 4


```{r Q4}
quiz(
  question("What is the odds ratio of the low dose vs the high dose?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 5

```{r Q5}
quiz(
  question("What is the standard error of the log odds ratio of the dose?",
    answer("approximately 0.5"),
    answer("approximately 1", correct=TRUE),
    answer("approximately 1.5"),
    answer("approximately 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 6

```{r Q6}
quiz(
  question("What is the upper limit of the 95% confidence interval of odds ratio of the dose? (high vs low)",
    answer("smaller than 0.5"),
    answer("between 0.5 and 1"),
    answer("between 1 and 1.5", correct=TRUE),
    answer("between 1.5 and 2"),
    answer("larger than 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

## Exercises (p1)

In this exercise we analyze the fracture data with logistic regression in R. This is a study concerning neurological deficit in thoracolumbar burst fractures. Between January 1981 and January 1991, 139 patients with thora-columbar or lumbar burst fractures were treated in the Department of Orthopaedics and the Department of Neurosurgery of the Academic Hospital Dijkzigt of the Erasmus University of Rotterdam (Fontijne WPJ et al., J.of Bone & Joint Surgery, in press).
A.o., the following variables were observed:


- `ID`: Subject number

- `DEF`: the occurrence of neurological deficit ('yes' or 'no')

- `LOC`: the localization of the fracture ('lumbar', 'thoracolumbar' or 'thoracic') 

- `STEN`: percentage spinal canal stenosis


f.	Now fit the model again after changing the coding of the dummies for LOC such that thoracic is used as reference. Before you run this model, try to find out what will be the estimates for the Constant and for the B’s of the dummies. When you run the new model Check whether the estimates are as expected.

g.	Next switch to the 'Deviation coding' contrast for LOC instead by running the command:
 contrasts(fracture$LOC) <- contr.sum(3)
In this case the effect of category 3 of LOC (the last category) is the negative sum of the effects of the categories 1 and 2.
Fit the model using this coding and compare the estimated coefficients again.

h.	What is the predicted probability of neurological deficit for a patient with a thoracic fracture and a stenosis of 60%?


i.	Create a scatter plot which shows how the predicted probability of deficit  depends on the localization and the percentage stenosis..


We will examine the effect of localization. This variable is actually an ordinal variable. It seems sensible to apply a trend test. This can be performed by including LOC as a continuous variable in the model.

j.	Fit a logistic model with LOC now as continuous variable (hint: use as.numeric) call this model glmNum.
What can you conclude now?
How is this related to what you have found in a.?

k.	Using the plot function on the fitted model gives you a number of diagnostic plots. Try to interpret the plots. 


l.	The functions cooks.distance and dfbetas gives you the cooks.distances and the dfbetas of the observations. Use:
 coo_1 <- cooks.distance(glmNum)
And try to find the patients with large value for Cook’s Leverage.
You may also sort the cases by the variable (in descending order) and print the cases with a high value. 
head(fracture[order(coo_1,decreasing = TRUE),], 10)



### Question 1


Display a couple of records from the data set to see what it looks like. You could use the `head` function. Make a cross table of the relation between the location of the fracture and the occurrence of neurological deficit. Is this relation significant?

```{r P1, exercise=TRUE, exercise.lines = 10, eval=FALSE}


```

```{r P1-hint-1, eval=FALSE}
# You could use the following functions
head(__)
tab3a <- table(__,__)
prop.table(___, ___)
chisq.test(___)
```
```{r P1-hint-2}
# Use:
head(fracture)
(tab3a <- table(fracture$LOC, fracture$DEF,deparse.level = 2))
prop.table(tab3a, margin = 1)   # columnwise perc
chisq.test(tab3a) # seems there is no relationship

```

### Question 2


Next look at the relationship between percentage stenosis and localization, describe and test it. How can you test which groups are significant different? 

```{r P2-setup}

```

```{r P2, exercise=TRUE, exercise.lines = 10, eval=FALSE}
# To describe the relation you can visualise it. For example by a boxplot. The 
# syntax is boxplot(outcome~group, data= dataset)
# Now estimate a logistic regression model using glm
# remember that the coef and confint can be used to get 
# the coefficients and their confidence intervals
# drop1 can be used to perform LR tests for all coefficients

```

```{r P2-hint-1}


```
```{r P2-hint-2}
# The code is:
boxplot(STEN~DEF, data=fracture)
# estimate the relationship using LOGISTIC regression
glm3b<-glm(DEF~STEN, family=binomial, data=fracture)
# Describe model by OR and CI
exp(coef(glm3b)) 
exp(confint(glm3b))
# test using a LR test
drop1(glm3b, test = 'LRT')  # very significant

# or do ttest
t.test(STEN~ DEF, data=fracture, var.equal = FALSE)
# also very significant

```

### Question 3

Next look at the relationship between percentage stenosis and localization, describe and test it. How can you test which groups are significant different?

```{r P3-setup}

```

```{r P3, exercise=TRUE, exercise.lines = 10}

```

```{r P3-hint-1, eval=FALSE}
# perform anova using lm
lm3c<-lm(___~ ___, data=fracture)
anova(___)
summary(___)

```


```{r P3-hint-2}
# The code is:
lm33<-lm(STEN~ LOC, data=fracture)
anova(lm33)
summary(lm33)
# better apply a bonferroni correction !!!
```

### Question 4

Now we will use logistic regression to examine the effect of percentage stenosis and localization on the probability of neurological deficit simultaneously.	Specify a model with the covariates LOC, STEN and the interaction between these two. Is it justified to fit a model without the interaction term?


```{r P4, exercise=TRUE, exercise.lines = 10, eval=FALSE}
drop1(___, test=___)

```

```{r P4-hint-1}
# Estimate a logistic regression model using glm
# use the drop1 or anova functions to perform a LR test 
```

```{r P4-hint-2}
# The code is:
glm34 <-glm(DEF~LOC*STEN,data=fracture, family = binomial)
summary(glm34)
exp(coef(glm34))
exp(confint(glm34))
# Lets see if we need interaction
drop1(glm34, test='LRT')
# OR
glm340 <-glm(DEF~LOC+STEN,data=fracture, family = binomial)
anova(glm34, glm340, test='LRT')

```


### Question 5

Now fit a logistic model without the interaction term. Save the predicted probabilities. We will use this variable later.
Looking at the results answer the following questions:
-	Does the percentage stenosis have a significant effect on neurological deficit, adjusted for localization? Describe the estimated effect.
-	Is localization significantly associated with the probability of neurological deficit (adjusted for percentage stenosis)? What is the P-value?
-	What are the odds ratios of the localization categories thoracic and thoracolumbar versus lumbar?
```{r P5-setup}

```

```{r P5, exercise=TRUE, exercise.lines = 10}


```


```{r P5-hint-1}
# The code is:
glm35 <-glm(DEF~LOC+STEN,data=fracture, family = binomial)
summary(glm35)
exp(coef(glm35)) 
exp(confint(glm35))
drop1(glm35, test='LRT')
# STENOSIS is significant 
# odds increase by 3% for each unit increase
# Location is not significant
# save predicted
preds <- predict(glm35,type = 'response')

```

### Question 6

Now fit the model again after changing the coding of the dummies for LOC such that thoracic is used as reference. In `glm` we can do this by including `c(LOC, base=3)` as a model term (3 because thoracic is the third level of this variable which you can check using the `levels` function). Before you run this model, try to find out what will be the estimates for the Constant and for the B’s of the dummies. When you run the new model Check whether the estimates are as expected.

```{r P6-setup}
```

```{r P6, exercise=TRUE, exercise.lines = 10, eval=FALSE}
glm36<- glm(~ C(LOC, base=3) + ___, ____)
exp(coef(glm36))

```

```{r P6-hint-1}

# Write down the equation of the model
# We know the how do the log odds change when we go change between the levels of 
# LOC. So how would they change if we go from thoracic to the other levels.
# knowing delta log odds allows us to compute the OR

```

```{r P6-hint-2}
# When LOC(thoracic) will be reference:
#  Constant: -1.373 - 1.307 = -2.680
# LOC(1) : lumbar/thracic = +1.307
# LOC(2) : thoracolumbar/thoracic = -0.353 - (-1.307) = 0.954
# Now take the exp
# The code is:
levels(fracture$LOC) # 3rd level is thoracic

glm36 <-glm(DEF~C(LOC, base=3)+STEN,data=fracture, family = binomial)
glm36
exp(coef(glm36))

```

### Question 7

Next switch to the 'Deviation coding' contrast for LOC instead by running the command:
 `contrasts(fracture$LOC) <- contr.sum(3)` 
In this case the effect of category 3 of LOC (the last category) is the negative sum of the effects of the categories 1 and 2.
Fit the model using this coding and compare the estimated coefficients again.
```{r P7-setup}

```

```{r P7, exercise=TRUE, exercise.lines = 10, eval=FALSE}
```

```{r P7-hint-1, eval=FALSE}
# Use the following template
contrastsbase<- contrasts(___)
contrasts(____ <- contr.sum(___)
glm37 <-glm(___)
glm37
exp(coef(glm37))
# restore contrast
contrasts(____)<-contrastsbase


```

```{r P7-hint-2}
# The code is:
contrastsbase<- contrasts(fracture$LOC)
contrasts(fracture$LOC) <- contr.sum(3)
glm37 <-glm(DEF~C(LOC, base=3)+STEN,data=fracture, family = binomial)
glm37
exp(coef(glm37))
# restore contrast
contrasts(fracture$LOC)<-contrastsbase

```



### Question 8

What is the predicted probability of neurological deficit for a patient with a thoracic fracture and a stenosis of 60%? (Using the moel from Ex 5.)

```{r P8-setup}
glm37 <-glm(DEF~C(LOC, base=3)+STEN,data=fracture, family = binomial)
glm35 <-glm(DEF~LOC+STEN,data=fracture, family = binomial)
```

```{r P8, exercise=TRUE, exercise.lines = 10, eval=FALSE}

```

```{r P8-hint-1}
# Use the predict function
```

```{r P8-hint-2}
# We can make a dataset for prediction with
preddata<-data.frame(LOC=factor('thoracic', levels = levels(fracture$LOC)), STEN=60)

```


```{r P8-hint-3}
# The code is:
preddata<-data.frame(LOC=factor('thoracic', levels = levels(fracture$LOC)), STEN=60)
predict(glm35, newdata = preddata, type='response' )

```

### Question 9

The functions cooks.distance and dfbetas gives you the cooks.distances and the dfbetas of the observations. Use:
` coo_1 <- cooks.distance(glm35)`
And try to find the patients with large value for Cook’s Leverage.
You may also sort the cases by the variable (in descending order) and print the cases with a high value. 
head(fracture[order(coo_1,decreasing = TRUE),], 10)

```{r P9-setup}

```

```{r P9, exercise=TRUE, exercise.lines = 12, eval=FALSE}

```

```{r P9-hint-1}
# use the code as isn the question
```

```{r P9-hint-2}
# The code is:
diagnost3<- data.frame(hat=hatvalues(glm35),dfbetas(glm35), cook=cooks.distance(glm35), fracture$ID)
head(diagnost3[order(diagnost3$cook,decreasing = TRUE),], 10)


```
