---
title: "Logistic regression 1"
output: 
  learnr::tutorial:
     allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
data("bmt", package = "EP03logistic")
knitr::opts_chunk$set(echo = FALSE)
```


## Quiz

The following questions test your knowledge about logistic regression, odds and odds ratios.

### Question 1

In Clements et. al. (1999) the effect of a high dose vs a low dose of D-Penicillamine in early diffuse systemic scleloris is investigated. 25 of the 36 patients that received a low dose responded while in the high dose group 17 out of 32 responded.
```{r Q1}
quiz(
  question("What is the probability of responding in the low dose group?",
    answer("Less than 40%"),
    answer("Between 40% and 60%"),
    answer("Between 60% and 80%", correct=TRUE),
    answer("More than 80%"),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 2


```{r Q2}
quiz(
  question("What are the odds of responding in the low dose group?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 3


```{r Q3}
quiz(
  question("What is the odds ratio of the low dose vs the high dose?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 4

```{r Q4}
quiz(
  question("What is the standard error of the log odds ratio of the dose?",
    answer("approximately 0.5"),
    answer("approximately 1", correct=TRUE),
    answer("approximately 1.5"),
    answer("approximately 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 5

```{r Q5}
quiz(
  question("What is the upper limit of the 95% confidence interval of odds ratio of the dose? (high vs low)",
    answer("< 0.5"),
    answer("between 0.5 and 1"),
    answer("between 1 and 1.5", correct=TRUE),
    answer("between 1.5 and 2"),
    answer("> 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

## Exercises

We will now examine how logistic regression can be used to investigate the risk of developing acute graft versus host disease after a bone marrow transplant. We will use the data set BMT.sav



### Question 1
Define a new variable sexmatch with value 1 if the receiver and donor have the same sex and value 0 if not. In R you can do this using:
library(foreign)
When reading in the SPSS file the file path between the quotes should correspond the location where the file is stored on your computer. R will give you a warning (as is frequently the case when importing SPSS files). In these exercises you can ignore this type of warning. 
b.	Make a cross table of the outcome agvhd against sexmatch using the table function. Add the proportions of healthy and deseased with the prop.table function. Use chisq.test to perform a 2-test.

Is there a significant relation between this variables?
Compute the odds ratio of the sexmatch = 1 versus sexmatch = 0. 

```{r P1, exercise=TRUE, exercise.lines = 10}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
(tab1b <- table(bmt$agvhd, bmt$sexmatch, deparse.level = 2))
prop.table(____, margin = 2)   
chisq.test(____)

```

```{r P1-hint-1}
# Just fill in the blanks
```
```{r P1-hint-2}
# Use:
(tab1b <- table(bmt$agvhd, bmt$sexmatch,deparse.level = 2))
prop.table(tab1b, margin = 2)   
chisq.test(tab1b)

```

### Question 2


Now we will look at the relation using logistic regression.

In R we can estimate a logistic regression model using the command `glm()`The first parameter indicates the model formula. On the left side of the ~ sign the dependent variable is specified (here agvhd); on the right hand side the explanatory variables (here the single variable sexmatch). The parameter family indicates the distribution we use. The default model for a binomial distribution is the logistic regression model. The last parameter specifies is the data set we use. 

```{r P2-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
```

```{r P2, exercise=TRUE, exercise.lines = 10}
glm1<-glm(___ ~ ____, family = ____, data=___)
glm1
```

```{r P2-hint-1}
# Just fill in the blanks

```
```{r P2-hint-2}
# The code is:
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)
glm1
```

### Question 3

Use the summary command on the returned object to obtain the estimated coefficients standard errors and p-values from a Wald test. The odds ratios can be obtained by using exp(coef(glm1)). The function confint is used to obtain confidence estimates, again the exp function transforms them to the odds-ratio scale. Compare the OR with the results obtained in b).


```{r P3-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)
 
```

```{r P3, exercise=TRUE, exercise.lines = 10}
```

```{r P3-hint-1}
# The code is:
# summary(___)
# exp(coef(___))
# exp(confint(___))

```


```{r P3-hint-2}
# The code is:
summary(glm1)
exp(coef(glm1))
exp(confint(glm1))

```

### Question 4


Using drop1(glm1, test='LR') we also obtain the p-values from a likelihood ratio (LR) test. Compare the results between the LR and Wald test. 

```{r P4-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)

```

```{r P4, exercise=TRUE, exercise.lines = 10}
drop1(___, test=___)

```

```{r P4-hint-1}
# Fill in the blanks
```

```{r P4-hint-2}
# The code is:
drop1(glm1, test='LRT')
```

