---
title: "Poisson regression"
output: 
  learnr::tutorial:
     allow_skip: true
tutorial:
  id: "io.shinyapps.erasmusmc.poisreg"
  version: 0.926     
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library(EP03logistic)
data("doll", package = "EP03logistic")
knitr::opts_chunk$set(echo = FALSE)
```


## Quiz

The following questions test your knowledge about Poisson regression

### Question 1
A trial is set up to compare a new procedure against usual care. We investigate the number of adverse events.
Consider the following output from a Poisson regression:

|              | beta     | Std. Error | z value |
|--------------|----------|------------|---------|
| (Intercept)  | -1.23782 | 0.42751    | -2.895  |
| age          | 0.08828  | 0.04947    | 1.785   |
| Intervention | 0.62459  | 0.31567    | 1.979   |



```{r Q1}
quiz(
  question("What is the p-value of the coefficient of age?",
    answer("smaller than 0.01"),
    answer("between 0.01 and 0.05"),
    answer("between 0.05 and 0.1", correct=TRUE),
    answer("0.1 or larger"),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 2


```{r Q2}
quiz(
  question("What is the p-value of the intervention?",
    answer("smaller than 0.01"),
    answer("between 0.01 and 0.05", correct=TRUE),
    answer("between 0.05 and 0.1"),
    answer("larger than 0.1"),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 3


```{r Q3}
quiz(
  question("What is the incidence rate ratio of the intervention?",
    answer("0.54"),
    answer("0.62"),
    answer("1.62"),
    answer("1.87", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 4

What is the lower limit of the 95% confidence interval of irr the intervention?
```{r Q4}
quiz(
  question(
"What is the lower limit of the 95% confidence interval of the intervention?", 
    answer("0.056"),
    answer("1.56"),
    answer("1.006", correct = TRUE),
    answer("2.006"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 5

What is the expected number of events of a twenty year old on the intervention group?
```{r Q5}
quiz(
  question(
"What is the lower limit of the 95% confidence interval of the intervention?", 
    answer("0.13"),
    answer("1.15"),
    answer("3.16", correct = TRUE),
    answer("9.3"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```


## Exercises

The table below presents dose-response data concerning the relationship between smoking and lung cancer taken from a publication by Doll. The data in the table were put in the data file DOLL.DAT. 

[Man-years at risk, number of cases of lung cancer (in parentheses)]

| **Cigarettes/day** |     **0**       |     **1-9**   |  **10-14**   |   **15-19**  |  **20-24**   |   **25-34**  |  **35 +**   |    *total*   |
|--------------|:-------------:|:-----------:|:----------:|:----------:|:----------:|:----------:|:---------:|:------------:|
|   *(mean)*    | (Non-smokers) |    (5.2)    |   (11.2)   |   (15.9)   |   (20.4)   |   (27.4)   |   (40.8)  |              |
| *Yrs of smoking*     |               |             |            |            |            |            |           |              |
| *15-19*        |   10366 (1)   |     3121    |    3577    |    4317    |    5683    |    3042    |    670    |   30776 (1)  |
| *20-24*        |      8162     |     2937    |  3286 (1)  |    4214    |  6385 (1)  |  4050 (1)  |    1166   |   30200 (3)  |
| *25-29*        |      5969     |     2288    |  2546 (1)  |    3185    |  5483 (1)  |  4290 (4)  |    1482   |   25243 (6)  |
| *30-34*        |      4496     |     2015    |  2219 (2)  |  2560 (4)  |  4687 (6)  |  4268 (9)  |  1580 (4) |  21825 (25)  |
| *35-39*        |      3512     |   1648 (1)  |    1826    |    1893    |  3646 (5)  |  3529 (9)  |  1336 (6) |  17390 (21)  |
| *40-44*        |      2201     |   1310 (2)  |  1386 (1)  |  1334 (2)  |  2411(12)  |  2424(11)  |  924(10)  |  11990 (38)  |
| *45-49*        |      1421     |     927     |   988 (2)  |   849 (2)  |  1567 (9)  |  1409(10)  |  556 (7)  |   7717 (30)  |
| *50-54*        |      1121     |   710 (3)   |   684 (4)  |   470 (2)  |   857 (7)  |   663 (5)  |  255 (4)  |   4760 (25)  |
| *55-59*        |    826 (2)    |     606     |   449 (3)  |   280 (5)  |   416 (7)  |   284 (3)  |  104 (1)  |   2965 (21)  |
| *total*        |  38074    (3) | 15562   (6) | 16961 (14) | 19102 (15) | 31135 (48) | 23959 (52) | 8073 (32) | 152866 (170) |


### Question 1

Confirm that the number of person years and the number of events in the data set `doll`. For this we need to weight the data. 

```{r P1-setup}
data(doll,package = 'EP03logistic')
```


```{r P1, exercise=TRUE, exercise.lines = 10, eval=FALSE}
#w1 <- rep((1:NROW(doll)),doll$prsyears)
#with(doll[w1,], table(___, ___))
#w2 <-
#with(____)  

```

```{r P1-hint-1}
# The weight we need to use are equal to the prsyears and death columns. We 
# need to repeat the row this number of times to use the table command 
```
```{r P1-hint-2}
# Use:
#w1 <- rep((1:NROW(doll)),doll$prsyears)
#with(doll[w1,], table(smoke, age))
#w2 <- rep((1:NROW(doll)),doll$death)
#with(doll[w2,], table(smoke, age))
# alternatively use
#xtabs(prsyears~age+smoke, data=doll)
#xtabs(death~age+smoke, data=doll)

```

### Question 2

Now we are going to estimate a Poisson model. First create the term to use as an offset variable. Now estimate a Poisson model using only an intercept (and the offset). Call the model glm1 What is the interpretation of the estimated coefficient? Also estimate the coefficient by hand using the table.



```{r P2-setup}
data(doll,package = 'EP03logistic')
```


```{r P2, exercise=TRUE, exercise.lines = 10}
#doll$logpy <- log(___)
#glm1<-glm(___~1+offset(____),  family=poisson, data=doll)
#summary(___)
```

```{r P2-hint-1}
# What should be the response? And which variable is the offset?
```
```{r P2-hint-2}
# The code is:
#doll$logpy <- log(doll$prsyears)
#glm1 <- glm(death~1+offset(logpy),  family=poisson, data=doll)
#summary(glm1)
```

### Question 3

Now fit the model with age category. Call the model glm2. Verify that you can calculate these coefficients also by hand.
What can you say about the goodness of fit.


```{r P3-setup}
data(doll,package = 'EP03logistic')
doll$logpy <- log(doll$prsyears)
```


```{r P3, exercise=TRUE, exercise.lines = 10}
glm1 <- glm(death~1+offset(logpy),  family=poisson, data=doll)
```

```{r P3-hint-1}
# Add the variables to the formula.  
```
```{r P3-hint-2}
# The code is:
#glm2 <- glm(death~1+age+offset(logpy),  family=poisson, data=doll)
#summary(glm2)
```

### Question 4

Now add the variable smoke. Call the model glm3. What can you say about the goodness of fit.



```{r P4-setup}

data(doll,package = 'EP03logistic')
doll$logpy <- log(doll$prsyears)
glm1 <- glm(death~1+offset(logpy),  family=poisson, data=doll)
glm2 <- glm(death~1+age+offset(logpy),  family=poisson, data=doll)
```

```{r P4, exercise=TRUE, exercise.lines = 10}

```

```{r P4-hint-1}
# Just add the variable to the formula
```

```{r P4-hint-2}
# The code is:
glm3 <- glm(death~1+as.factor(age)+smoke+offset(logpy),  family=poisson, data=doll)
summary(glm3)
anova(glm3, glm2, test = 'LRT')
```

### Question 5
Now we check if the effect of age can be better modlled by a linear trend. Use `as.numeric` to convert the factor to a numeric value. Call the model glm4. Does this model do better than the previous model (use a likelihood ratio test)?


```{r P5-setup}
data(doll,package = 'EP03logistic')
doll$logpy <- log(doll$prsyears)
glm1 <- glm(death~1+offset(logpy),  family=poisson, data=doll)
glm2 <- glm(death~1+age+offset(logpy),  family=poisson, data=doll)
glm3 <- glm(death~1+age+smoke+offset(logpy),  family=poisson, data=doll)


```



```{r P5, exercise=TRUE, exercise.lines = 10}



```






```{r P5-hint-1}
# substitute age by as.numeric(age) in glm3
```
```{r P5-hint-2}
# the code is
doll$numage <- as.numeric(doll$age)
glm4 <- glm(death~1+numage+smoke+offset(logpy),  family=poisson, data=doll)
anova(glm4, glm3,test = 'LRT')
```


### Question 6
Now use  `log(as.numeric(age))` to model the effect of age. What about the goodness-of-fit? Does this do better than the previous model?

```{r P6, exercise=TRUE, exercise.lines = 10}

```

```{r P6-setup}
doll$logpy <- log(doll$prsyears)
glm1 <- glm(death~1+offset(logpy),  family=poisson, data=doll)
glm2 <- glm(death~1+age+offset(logpy),  family=poisson, data=doll)
glm3 <- glm(death~1+age+smoke+offset(logpy),  family=poisson, data=doll)
glm4 <- glm(death~1+as.numeric(age)+smoke+offset(logpy),  family=poisson, data=doll)
doll$numage <- as.numeric(doll$age)
```



```{r P6-hint-1}
# If models are not nested you cannot use a LRT. Let's use AIC instead.
```
```{r P6-hint-2}
# the code is
glm5 <- glm(death~1+I(log(numage))+smoke+offset(logpy),  family=poisson, data=doll)

AIC(glm1, glm2, glm3, glm4, glm5)
```
