---
title: "Logistic regression 1"
output: 
  learnr::tutorial:
     allow_skip: true
tutorial:
  id: "io.shinyapps.erasmusmc.logisticregression1"
  version: 0.921
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library('EP03logistic')
data("bmt", package = "EP03logistic")
knitr::opts_chunk$set(echo = FALSE)
```


## Quiz

The following questions test your knowledge about logistic regression, odds and odds ratios.

### Question 1
```{r Q1}
quiz(
  question("True or false (check all that apply)", 
    type='multiple',         
    answer("The confidence interval around an odds ratio is symetric"),
    answer("The log of the coefficients of an logistic regression can be interpreted as odds ratios"),
    answer("An odds of one correspons to a probability of 50%", correct=TRUE),
    answer("When a probability is one, the corresponding odds are infinite", correct=TRUE),
    answer("The logistic function cannot reach values larger than one", correct=TRUE),
    answer("The logit function can be used to convert probabilities to odds ", correct=TRUE),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 2
```{r Q2}
quiz(
  question("True or false (check all that apply)", 
    type='multiple',         
    answer("The logistic regression model can only be used with continuous independant variables"),
    answer("The logistic regression model can only be used with continuous dependant variables"),
    answer("The logistic regression model can only be used with binary independant variables"),
    answer("The logistic regression model can only be used with binary dependant variables", correct=TRUE),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 3

In Clements et. al. (1999) the effect of a high dose vs a low dose of D-Penicillamine in early diffuse systemic scleloris is investigated. 25 of the 36 patients that received a low dose responded while in the high dose group 17 out of 32 responded.
```{r Q3}
quiz(
  question("What is the probability of responding in the low dose group?",
    answer("Less than 40%"),
    answer("Between 40% and 60%"),
    answer("Between 60% and 80%", correct=TRUE),
    answer("More than 80%"),
        allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 4


```{r Q4}
quiz(
  question("What are the odds of responding in the low dose group?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 5


```{r Q5}
quiz(
  question("What is the odds ratio of the low dose vs the high dose?",
    answer("Less than 0.5"),
    answer("Between 0.5 and 1"),
    answer("Between 1 and 2"),
    answer("More than 2", correct=TRUE),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 6

```{r Q6}
quiz(
  question("What is the standard error of the log odds ratio of the dose?",
    answer("approximately 0.5"),
    answer("approximately 1", correct=TRUE),
    answer("approximately 1.5"),
    answer("approximately 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 7

```{r Q7}
quiz(
  question("What is the upper limit of the 95% confidence interval of odds ratio of the dose? (high vs low)",
    answer("smaller than 0.5"),
    answer("between 0.5 and 1"),
    answer("between 1 and 1.5", correct=TRUE),
    answer("between 1.5 and 2"),
    answer("larger than 2"),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

## Exercises (p1)

We will now examine how logistic regression can be used to investigate the risk of developing acute graft versus host disease after a bone marrow transplant. We will use the data set BMT.sav



### Question 1
Define a new variable sexmatch with value 1 if the receiver and donor have the same sex and value 0 if not. In R you can do this using:
library(foreign)
When reading in the SPSS file the file path between the quotes should correspond the location where the file is stored on your computer. R will give you a warning (as is frequently the case when importing SPSS files). In these exercises you can ignore this type of warning. 
b.	Make a cross table of the outcome agvhd against sexmatch using the table function. Add the proportions of healthy and deseased with the prop.table function. Use chisq.test to perform a 2-test.

Is there a significant relation between this variables?
Compute the odds ratio of the sexmatch = 1 versus sexmatch = 0. 

```{r P1, exercise=TRUE, exercise.lines = 10, eval=FALSE}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
(tab1b <- table(bmt$agvhd, bmt$sexmatch, deparse.level = 2))
prop.table(____, margin = 2)   
chisq.test(____)

```

```{r P1-hint-1}
# Just fill in the blanks
```
```{r P1-hint-2}
# Use:
(tab1b <- table(bmt$agvhd, bmt$sexmatch,deparse.level = 2))
prop.table(tab1b, margin = 2)   
chisq.test(tab1b)

```

### Question 2


Now we will look at the relation using logistic regression.

In R we can estimate a logistic regression model using the command `glm()`The first parameter indicates the model formula. On the left side of the ~ sign the dependent variable is specified (here agvhd); on the right hand side the explanatory variables (here the single variable sexmatch). The parameter family indicates the distribution we use. The default model for a binomial distribution is the logistic regression model. The last parameter specifies is the data set we use. 

```{r P2-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
```

```{r P2, exercise=TRUE, exercise.lines = 10, eval=FALSE}
glm1<-glm(___ ~ ____, family = ____, data=___)
glm1
```

```{r P2-hint-1}
# Just fill in the blanks

```
```{r P2-hint-2}
# The code is:
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)
glm1
```

### Question 3

Use the summary command on the returned object to obtain the estimated coefficients standard errors and p-values from a Wald test. The odds ratios can be obtained by using exp(coef(glm1)). The function confint is used to obtain confidence estimates, again the exp function transforms them to the odds-ratio scale. Compare the OR with the results obtained in b).


```{r P3-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)
 
```

```{r P3, exercise=TRUE, exercise.lines = 10}
```

```{r P3-hint-1}
# The code is:
# summary(___)
# exp(coef(___))
# exp(confint(___))

```


```{r P3-hint-2}
# The code is:
summary(glm1)
exp(coef(glm1))
exp(confint(glm1))

```

### Question 4


Using drop1(glm1, test='LR') we also obtain the p-values from a likelihood ratio (LR) test. Compare the results between the LR and Wald test. 

```{r P4-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glm1<-glm(agvhd ~ sexmatch, family = 'binomial', data=bmt)

```

```{r P4, exercise=TRUE, exercise.lines = 10, eval=FALSE}
drop1(___, test=___)

```

```{r P4-hint-1}
# Fill in the blanks
```

```{r P4-hint-2}
# The code is:
drop1(glm1, test='LRT')
```

## Exercises (p2)

We will now look at the effect of a continous variable on the odds of agvhd. 	

### Question 5

Specify a model using agedon as covariate. What is the estimated odds ratio of this continuous variable and the confidence interval? What is the meaning of the odds ratio? Call the estimated model glmAge.

```{r P5-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
```

```{r P5, exercise=TRUE, exercise.lines = 10}


```

```{r P5-hint-1}
# Use the glm function as before.
```

```{r P5-hint-2}
# The code is:
glmAge<-glm(agvhd ~ agedon, family = 'binomial', data=bmt)
exp(coef(glmAge))
exp(confint(glmAge))
```

### Question 6

We are now going to plot the relation between the age of the donor and the estimated probabilities. We can use the `predict` function for this. The first argument is  the estimated `glm`. When we use `type = 'response'` we  obtain the predictions on the scale of the response (the 0-1 scale of probabilities, this in contrast to the scale of the linear predictor).

```{r P6-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glmAge<-glm(agvhd ~ agedon, family = 'binomial', data=bmt)
```

```{r P6, exercise=TRUE, exercise.lines = 10, eval=FALSE}
y <- predict(____, type = ____)

plot(x=bmt$agedon[order(bmt$agedon)],  
  y=y[order(bmt$agedon)],
  type='l')
```

```{r P6-hint-1}
# Fill in the name of the model
```

```{r P6-hint-2}
# The code is:
y <- predict(glmAge, type = 'response')
plot(x=bmt$agedon[order(bmt$agedon)],  
  y=y[order(bmt$agedon)],
  type='l')
```

### Question 7

In the model above we assume a linear relation between the log odds for agedon and the outcome. To test this assumption we can add the square of agedon to the model. Add the quadratic term to the model below (hint: use I(agedon^2)). Call the model glmAge2. Is the quadratic term significant?
```{r P7-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glmAge<-glm(agvhd ~ agedon, family = 'binomial', data=bmt)
```

```{r P7, exercise=TRUE, exercise.lines = 10, eval=FALSE}
glmAge2<-glm(____ ~ agedon+___, _____)
anova(glmAge2, glmAge, test='LRT')
```

```{r P7-hint-1}
# Fill in the response and the quadratic term as specified above
```

```{r P7-hint-2}
# The code is:
glmAge<-glm(agvhd ~ agedon+I(agedon^2), family = 'binomial', data=bmt)
```

## Exercises (p3)

Now we will build a multivariable model using multiple explanitory variables.

### Question 8

Estimate a model using `agedon`, `agedon` squared, `agerec`, `sexmatch` and `diag`. What are the odds ratios?


```{r P8-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glmAge<-glm(agvhd ~ agedon, family = 'binomial', data=bmt)

```

```{r P8, exercise=TRUE, exercise.lines = 10, eval=FALSE}
glmMV <-

```

```{r P8-hint-1}
# Use the glm function as before
```

```{r P8-hint-2}
# The code is:
glmMV<-glm(agvhd ~ agedon+I(agedon^2)+agerec+sexmatch+diag, family = 'binomial', data=bmt)
exp(coef(glmMV))
exp(confint(glmMV))
```

### Question 9

Let’s visualize the risk of acute graft versus host desease for a 30 year old male  with acute nonlymphoblastic leukemia with a male and a female donor.

```{r P9-setup}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
glmAge<-glm(agvhd ~ agedon, family = 'binomial', data=bmt)
glmMV<-glm(agvhd ~ agedon+I(agedon^2)+agerec+sexmatch+diag, family = 'binomial', data=bmt)

```

```{r P9, exercise=TRUE, exercise.lines = 12, eval=FALSE}
predict_data <- expand.grid(agedon=0:60,
           diag='acute nonlymphoblastic leukemia', 
           sexdon=levels(bmt$sexdon),
           sexrec=____, agerec=____)
predict_data$sexmatch <- ifelse(predict_data$sexdon=='female',0,1) 
predict_data$y <- predict(______)
plot(x=___, y=____, type='p',
     col=ifelse(predict_data$sexdon=='male', 'blue', 'hotpink'))

```

```{r P9-hint-1}
# Fill in the blanks
```

```{r P9-hint-2}
# The code is:
predict_data <- expand.grid(agedon=0:60,
           diag='acute nonlymphoblastic leukemia', 
           sexdon=levels(bmt$sexdon),
           sexrec='male', agerec=30)
predict_data$sexmatch <- ifelse(predict_data$sexdon=='female',0,1) 
predict_data$y <- predict(glmMV, type = 'response',
  newdata=predict_data)
plot(x=predict_data$agedon, y=predict_data$y, type='p',
     col=ifelse(predict_data$sexdon=='male', 'blue', 'hotpink'))
```
