---
title: "Logistic regression 1"
output: 
  # learnr::tutorial:
  #   allow_skip: true
 pdf_document
#tutorial:
#  id: "io.shinyapps.erasmusmc.logisticregression1"
#  version: 0.923
#runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library('EP03logistic')
data("bmt", package = "EP03logistic")
knitr::opts_chunk$set(echo = FALSE)
```



## Question 1

We will examine how logistic regression can be used to investigate the risk of developing acute graft versus host disease after a bone marrow transplant. We will use the data set BMT.sav. In R you can import the data set using the commands:
```{r, eval=FALSE}
library(foreign)
bmt <- read.spss("/path/to/file/BMT.sav", to.data.frame = TRUE)
```
When reading in the SPSS file, the file path between the quotes should correspond the location where the file is stored on your computer (remember to use forward slashes instead of backslashes to seperate folders). R will give you a warning (as is frequently the case when importing SPSS files). In these exercises you can ignore this type of warning. 
Here (in the shiny app), we have allready loaded the bmt data.

### Question 1

a) Define a new variable sexmatch with value 1 if the receiver and donor have the same sex and value 0 if not. In R you can do this using:
```{r, eval=FALSE}
bmt$sexmatch <- ifelse(bmt$sexrec==bmt$sexdon, 1, 0)
```
b.	Make a cross table of the outcome agvhd against sexmatch using the `table` function. Add the proportions of healthy and deseased with the prop.table function. Is there a significant relation between this variables? Use `chisq.test` or `prop.test` to perform a chi2-test.


### Question 2
Compute the odds ratio of the sexmatch = 1 versus sexmatch = 0. Use R as a calculator to calculate the odds ratio from the crosstabulation table.

### Question 3

Now we will look at the relation using logistic regression.

In R we can estimate a logistic regression model using the command `glm()`The first parameter indicates the model formula. On the left side of the ~ sign the dependent variable is specified (here `agvhd`); on the right hand side the explanatory variables (here the single variable sexmatch). The parameter family indicates the distribution we use. The default model for a binomial distribution is the logistic regression model. The last parameter specifies  the data set we use. 

### Question 4

Use the summary command on the returned object to obtain the estimated coefficients standard errors and p-values from a Wald test. The odds ratios can be obtained by using exp(coef(glm1)). The function confint is used to obtain confidence estimates, again the exp function transforms them to the odds-ratio scale. Compare the OR with the result you obtained when you calculated it by hand.

### Question 5

Using `drop1(glm1, test='LRT')` we also obtain the p-values from a likelihood ratio (LR) test. Compare the results between the LR and Wald test. 


We will now look at the effect of a continous variable on the odds of `agvhd`. 	

### Question 6

Specify a model using `agedon` as covariate. What is the estimated odds ratio of this continuous variable and the confidence interval? What is the meaning of the odds ratio? 


### Question 7

We are now going to plot the relation between the age of the donor and the estimated probabilities. We can use the `predict` function for this. The first argument is  the estimated `glm`. When we use `type = 'response'` we  obtain the predictions on the scale of the response (the 0-1 scale of probabilities, this in contrast to the scale of the linear predictor).

### Question 8

In the model above we assume a linear relation between the log odds for agedon and the outcome. To test this assumption we can add the square of agedon to the model. Add a quadratic term to the model (hint: use `I(agedon^2)`).  Is the quadratic term significant?

### Question 9

Estimate a model using `agedon`, `agedon` squared, `agerec`, `sexmatch` and `diag` as predictors. What are the odds ratios?


### Question 10

Let's visualize the risk of acute graft versus host desease for a 30 year old male  with acute nonlymphoblastic leukemia with a male and a female donor. Use the following code as a template.


```{r P9, exercise=TRUE, exercise.lines = 12, eval=FALSE}
predict_data <- expand.grid(agedon=0:60,
           diag='acute nonlymphoblastic leukemia', 
           sexdon=levels(bmt$sexdon),
           sexrec=____, agerec=____)
predict_data$sexmatch <- ifelse(predict_data$sexdon=='female',0,1) 
predict_data$y <- predict(______)
plot(x=___, y=____, type='p',
     col=ifelse(predict_data$sexdon=='male', 'blue', 'hotpink'))

```

